HO unification from object language to meta language

### 2024-04-05 paper/main.tex
typos

\begin{abstract}
  Specifying and implementing a logic from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are simplified when ML binders
  represent object logic ones; 2) proof construction, and even proof search, is
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
  Constuctions~\cite{felty93lics}.

@@

Specifying and implementing a logic from scratch requires significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways: 1) variable binding and substitution are simplified when ML binders
represent object logic ones; 2) proof construction, and even proof search, is
greatly simplified by leveraging the unification procedure provided by the ML.
Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
which have been utilized to implement various formal systems such as
First Order Logic~\cite{felty88cade},
Set Theory~\cite{10.1007/BF00881873},
Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
Constuctions~\cite{felty93lics}.

@@

XXX permuting hrun does not change the final result if check dooes not fail eagerly

@@

\noindent
New links are suspended by construction.
If the left hand side variable is assigned during a step, then 
the link is considered for progress and possibly eliminated.
This is discussed in \cref{sec:eta} and \cref{sc:beta}.

@@

\section[Compilation: fo\_tm to tm]{Basic simulation of \Fo{} in \Ho{}}
\label{sec:compilation}

@@

\noindent
Decompiling an assignment requires to turn abstractions into
lambdas. For aestetic purposes we also eta-contract the result
(not needed since \Fo{} equality can do that)

@@

% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
\noteD[inline]{I've rewritten it, it is clearer?}
Until now, we have only dealt we unification of terms in \llambda. However, we
want the unification relation to be more robust so that it can work with terms
in \maybebeta. In general, unification in \maybebeta admits more then one
solution and commiting one of them in the substitution does not guarantee
prop.~\ref{prop:complete}. For instance, $X \appsep a \Uo a$ is a unification
problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
future unifications.

@@

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$, can be solved in \llambda.
\noteE[inline]{it is even a ground term, there is no unification left to perform actually}
\noteD[inline]{i don't understand the note}
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifable problem $(\lambda x. B) \appsep a \Ue a$.

To address
this unification problem, term compilation should capture the
terms $t$ in \maybebeta and replace them with fresh varaibles $X$.
The variables $X$ and the terms $t$ are linked through a \linkbeta.

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH, then it desappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates.
  \noteD{L'ho riscritto}

  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta desappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

@@

\begin{proof}[Proof sketch]
  If $\sigma\hoUnifPb$ is in \llambda, then by
  \cref{def:progBetaLL,def:progBetaRH}, then \linkbeta desappear and the
  unification done between terms in \llambda. This problem unifies iff
  its corresponding \Fo{} problem unifies too. If $\sigma\hoUnifPb$ is in \maybebeta,
  then, by \cref{lemma:keep-fidelity}, the unification fails, as per the
  corresponding unification in \Fo{}.
\end{proof}

@@

\noindent
Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
The first unification problem results in $\sigma =
\{A \mapsto \lambda x.x\}$. In turn this instantiation
triggers $\linkStore_1$ by \progressetaleft 
and $E$ is assigned to $\lambda x.x$.
Under this substitution the \linkbeta becomes
\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
it is replaced with the link:
\linketaM{}{E}{\lambda x.D_x}, while $C_x$ is unified with $D_x$. The second unification
problem assignes $f$ to $B$, that in turn activates
the second \linketa ($f$ is assigned to $C$), and then all the remaining links
are solved. The final \Ho substitution is $\sigma = \{A _{} \mapsto \lambda x.x, 
B _{} \mapsto f, C _{x} \mapsto (f\appsep x), 
D _{} \mapsto f, E _{x} \mapsto x, F _{x} \mapsto C_{x} \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
Y \mapsto f\}$.
\noteE{alla fine non si usa anche la regola beta-rigid-head?}
\noteD{nope. Si risolve tutto by \progBetaLL}

@@

In the example above, $\foUnifPb_1$ is in \maybebeta and the object language
cannot solve it, and, by \cref{prop:fidelity}, the meta language neither.
However, we can be more permissive, and relax \cref{lemma:keep-fidelity}. This
modification is quite simple to manage: we are introducing a new \maybebeta
progress rule, say \progBetaNoLLWait, by which, if \lhs is rigid and \rhs is
flexible, the considered \linkbeta is kept in the store and no progression is
done\footnote{This new rule trivially guarantees the termination of progress}.
\progBetaNoLLWait makes \elpiIn{occur-check-links} partial, since the check is
possible only on links with a variable on the \lhs. This means that we can have
two links \linkbetaM{}{X}{Y\appsep a} and \linkbetaM{}{f\appsep X}{Y\appsep a}
where the occur check does not thow an error. Note however, that the
decompilation of the two links will force the unification of $X$ to $Y\appsep a$
and then the unification of $f\appsep (Y \appsep a)$ to $Y\appsep a$, which
fails by the occur check of \Ue.

>>>>

\begin{abstract}
  Specifying and implementing a logic from scratch requires significant effort.
  Logical Frameworks and Higher Order Logic Programming Languages provide
  dedicated, high-level Meta Languages (ML) to facilitate this task in two
  key ways: 1) variable binding and substitution are simplified when ML binders
  represent object logic ones; 2) proof construction, and even proof search, is
  greatly simplified by leveraging the unification procedure provided by the ML.
  Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
  $\lambda$Prolog~\cite{miller_nadathur_2012} and
  Isabelle~\cite{10.1007/978-3-540-71067-7_7}
  which have been utilized to implement various formal systems such as
  First Order Logic~\cite{felty88cade},
  Set Theory~\cite{10.1007/BF00881873},
  Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
  Constructions~\cite{felty93lics}.

@@

Specifying and implementing a logic from scratch requires significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways: 1) variable binding and substitution are simplified when ML binders
represent object logic ones; 2) proof construction, and even proof search, is
greatly simplified by leveraging the unification procedure provided by the ML.
Notable examples of ML are Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
which have been utilized to implement various formal systems such as
First Order Logic~\cite{felty88cade},
Set Theory~\cite{10.1007/BF00881873},
Higher Order Logic~\cite{books/sp/NipkowPW02}, and even the Calculus of
Constructions~\cite{felty93lics}.

@@

XXX permuting hrun does not change the final result if check does not fail eagerly

@@

\noindent
New links are suspended by construction.
If the left hand side variable is assigned during a step, then 
the link is considered for progress and possibly eliminated.
This is discussed in \cref{sec:eta} and \cref{sec:beta}.

@@

\section[Compilation: fo\_tm to tm]{Basic simulation of \Fo{} in \Ho{}}
% \label{sec:compilation}

@@

\noindent
Decompiling an assignment requires to turn abstractions into
lambdas. For aesthetic purposes we also eta-contract the result
(not needed since \Fo{} equality can do that)

@@

% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
\noteD[inline]{I've rewritten it, it is clearer?}
Until now, we have only dealt we unification of terms in \llambda. However, we
want the unification relation to be more robust so that it can work with terms
in \maybebeta. In general, unification in \maybebeta admits more then one
solution and committing one of them in the substitution does not guarantee
prop.~\ref{prop:complete}. For instance, $X \appsep a \Uo a$ is a unification
problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
future unifications.

@@

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$, can be solved in \llambda.
\noteE[inline]{it is even a ground term, there is no unification left to perform actually}
\noteD[inline]{i don't understand the note}
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifiable problem $(\lambda x. B) \appsep a \Ue a$.

To address
this unification problem, term compilation should capture the
terms $t$ in \maybebeta and replace them with fresh variables $X$.
The variables $X$ and the terms $t$ are linked through a \linkbeta.

@@

\begin{proof}[Proof sketch]
  Let $l$ a \linkbeta in the store \linkStore. If $l$ is activated by
  \progBetaRH, then it disappears from \linkStore and \elpiIn{progress}
  terminates. Otherwise, the \rhs of $l$ is made by a variable applied to $m$
  arguments. At each activation of \progBetaLL, $l$ is replaced by a new
  \linkbeta $l^1$ having $m-1$ arguments. At the $m^{th}$ iteration, the
  \linkbeta $l^m$ has no more arguments and is removed from \linkStore.
  Note that at the $m^{th}$ iteration, $m$ new \linketa have been added to
  \linkStore, however, by \cref{lemma:prog-eta-terminates}, the algorithm
  terminates.
  \noteD{L'ho riscritto}

  % \Cref{def:progBetaRH} makes \elpiIn{progress} terminates, since it makes a
  % \linkbeta disappear from \linkStore. On the other hand, \cref{def:progBetaLL},
  % creates each time a new \linketa and replace the old \linkbeta with a new one.
  % This progression can however triggered at most $m$ times, since each new
  % \linkbeta as a smaller list of applied variables. At the $m^{th}$ progression,
  % the \linkbeta is removed from \linkStore which is now filled by $m$ new
  % \linketa. By \cref{lemma:prog-eta-terminates}, we know that \elpiIn{progress}
  % terminates if \linkStore is made by only \linketa and therefore
  % \elpiIn{progress} still terminates.
  %\noteE[inline]{is it ok?}
  \noteE[inline]{funziona. per essere più precisi io parlerei di ordine lessicografico (tipico ordine ben fondato usato per dimostrare terminazione). Nl nostro caso è la tripla (argomenti extra dei beta, numero di beta, numero di eta).}
\end{proof}

@@

\begin{proof}[Proof sketch]
  If $\sigma\hoUnifPb$ is in \llambda, then by
  \cref{def:progBetaLL,def:progBetaRH}, then \linkbeta disappear and the
  unification done between terms in \llambda. This problem unifies iff
  its corresponding \Fo{} problem unifies too. If $\sigma\hoUnifPb$ is in \maybebeta,
  then, by \cref{lemma:keep-fidelity}, the unification fails, as per the
  corresponding unification in \Fo{}.
\end{proof}

@@

\noindent
Initially the \linkbeta \rhs is a variable $D$ applied to the $E_x$.
The first unification problem results in $\sigma =
\{A \mapsto \lambda x.x\}$. In turn this instantiation
triggers $\linkStore_1$ by \progressetaleft 
and $E$ is assigned to $\lambda x.x$.
Under this substitution the \linkbeta becomes
\linkbetaM{x}{C_x}{(D \appsep x)}, and by \progBetaLL
it is replaced with the link:
\linketaM{}{E}{\lambda x.D_x}, while $C_x$ is unified with $D_x$. The second unification
problem assigns $f$ to $B$, that in turn activates
the second \linketa ($f$ is assigned to $C$), and then all the remaining links
are solved. The final \Ho substitution is $\sigma = \{A _{} \mapsto \lambda x.x, 
B _{} \mapsto f, C _{x} \mapsto (f\appsep x), 
D _{} \mapsto f, E _{x} \mapsto x, F _{x} \mapsto C_{x} \}$ and is decompiled into $\rho = \{X \mapsto \lambda x.x,
Y \mapsto f\}$.
\noteE{alla fine non si usa anche la regola beta-rigid-head?}
\noteD{nope. Si risolve tutto by \progBetaLL}

@@

In the example above, $\foUnifPb_1$ is in \maybebeta and the object language
cannot solve it, and, by \cref{prop:fidelity}, the meta language neither.
However, we can be more permissive, and relax \cref{lemma:keep-fidelity}. This
modification is quite simple to manage: we are introducing a new \maybebeta
progress rule, say \progBetaNoLLWait, by which, if \lhs is rigid and \rhs is
flexible, the considered \linkbeta is kept in the store and no progression is
done\footnote{This new rule trivially guarantees the termination of progress}.
\progBetaNoLLWait makes \elpiIn{occur-check-links} partial, since the check is
possible only on links with a variable on the \lhs. This means that we can have
two links \linkbetaM{}{X}{Y\appsep a} and \linkbetaM{}{f\appsep X}{Y\appsep a}
where the occur check does not throw an error. Note however, that the
decompilation of the two links will force the unification of $X$ to $Y\appsep a$
and then the unification of $f\appsep (Y \appsep a)$ to $Y\appsep a$, which
fails by the occur check of \Ue.
